---
title: "Diabetes Readmission Predictive Modelling"
author: "Manasvi Saite"
output: html_document
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(dplyr)
select <- dplyr::select
library(caret)
library(randomForest)
library(pROC)
```

## Load Data & Data Cleaning

```{r}
diabetes <- read.csv("diabetic_data.csv", stringsAsFactors = FALSE)

str(diabetes)
dim(diabetes)
summary(diabetes)

table(diabetes$readmitted)

diabetes[diabetes == "?"] <- NA

diabetes$readmit30 <- ifelse(diabetes$readmitted == "<30", 1, 0)
diabetes$readmit30 <- factor(diabetes$readmit30, levels = c(0,1), labels = c("No", "Yes"))

table(diabetes$readmit30)

diabetes <- diabetes %>%
  select(-encounter_id, -patient_nbr, -readmitted)

missing_summary <- sapply(diabetes, function(x) sum(is.na(x)))
missing_table <- data.frame(
  Variable = names(missing_summary),
  Missing_Count = as.numeric(missing_summary),
  Missing_Percent = round(as.numeric(missing_summary) / nrow(diabetes) * 100, 2)
)
missing_table <- missing_table[order(-missing_table$Missing_Count), ]
head(missing_table, 15)
```

## Choose predictors and make a smaller working dataset

```{r}
model_data <- diabetes %>%
  select(readmit30, race, gender, age, admission_type_id,
         discharge_disposition_id, admission_source_id,
         time_in_hospital, num_lab_procedures, num_procedures,
         num_medications, number_outpatient, number_emergency,
         number_inpatient, number_diagnoses, max_glu_serum,
         A1Cresult, insulin, change, diabetesMed)

str(model_data)
summary(model_data)
```

## Convert categorical variables

```{r}
model_data$readmit30 <- factor(model_data$readmit30, levels = c("No", "Yes"))
model_data$race <- factor(model_data$race)
model_data$gender <- factor(model_data$gender)
model_data$age <- factor(model_data$age, ordered = TRUE)
model_data$admission_type_id <- factor(model_data$admission_type_id)
model_data$discharge_disposition_id <- factor(model_data$discharge_disposition_id)
model_data$admission_source_id <- factor(model_data$admission_source_id)
model_data$max_glu_serum <- factor(model_data$max_glu_serum)
model_data$A1Cresult <- factor(model_data$A1Cresult)
model_data$insulin <- factor(model_data$insulin)
model_data$change <- factor(model_data$change)
model_data$diabetesMed <- factor(model_data$diabetesMed)

str(model_data)
```

## Remove missing values for first simple model

```{r}
model_data_clean <- na.omit(model_data)
dim(model_data_clean)
prop.table(table(model_data_clean$readmit30))
```

## Take a random sample to make modelling more manageable

```{r}
set.seed(123)

sample_n_rows <- min(20000, nrow(model_data_clean))
model_data_sample <- model_data_clean[sample(seq_len(nrow(model_data_clean)), sample_n_rows), ]

dim(model_data_sample)
prop.table(table(model_data_sample$readmit30))
```

## Train-test split

```{r}
set.seed(123)

train_index <- createDataPartition(model_data_sample$readmit30, p = 0.7, list = FALSE)
train_data <- model_data_sample[train_index, ]
test_data  <- model_data_sample[-train_index, ]

dim(train_data)
dim(test_data)
```

## Fix factor levels and variable types in test data

```{r}
factor_vars <- names(train_data)[sapply(train_data, is.factor)]

keep_rows <- rep(TRUE, nrow(test_data))

for (v in factor_vars) {
  train_levels <- unique(as.character(train_data[[v]]))
  test_values  <- as.character(test_data[[v]])
  keep_rows <- keep_rows & (is.na(test_values) | test_values %in% train_levels)
}

test_data <- test_data[keep_rows, ]

for (v in names(train_data)) {
  if (is.factor(train_data[[v]])) {
    test_data[[v]] <- factor(
      test_data[[v]],
      levels = levels(train_data[[v]]),
      ordered = is.ordered(train_data[[v]])
    )
  } else if (is.integer(train_data[[v]])) {
    test_data[[v]] <- as.integer(test_data[[v]])
  } else if (is.numeric(train_data[[v]])) {
    test_data[[v]] <- as.numeric(test_data[[v]])
  } else if (is.character(train_data[[v]])) {
    test_data[[v]] <- as.character(test_data[[v]])
  }
}

train_data <- droplevels(train_data)
test_data  <- droplevels(test_data)

str(train_data)
str(test_data)
```

## Logistic regression

```{r}
log_model <- glm(readmit30 ~ ., data = train_data, family = binomial)

summary(log_model)
```

## Odds ratios from logistic regression

```{r}
log_or <- data.frame(
  Variable = names(coef(log_model)),
  Odds_Ratio = exp(coef(log_model)),
  row.names = NULL
)
head(log_or, 15)
```

## Logistic regression prediction

```{r}
log_probs <- predict(log_model, newdata = test_data, type = "response")
log_pred <- ifelse(log_probs > 0.5, "Yes", "No")
log_pred <- factor(log_pred, levels = c("No", "Yes"))

log_cm <- confusionMatrix(log_pred, test_data$readmit30, positive = "Yes")
log_cm
```

## Logistic regression ROC-AUC

```{r}
log_roc <- roc(test_data$readmit30, log_probs, 
               levels = c("No", "Yes"), direction = "<")

log_auc <- auc(log_roc)
log_auc

plot(log_roc, col = "pink", lwd = 2,
     main = "ROC Curve - Logistic Regression")
```

## Random forest

```{r}
set.seed(123)

rf_model <- randomForest(readmit30 ~ ., data = train_data, ntree = 100, importance = TRUE)

print(rf_model)
varImpPlot(rf_model)
```

## Random forest prediction

```{r}
for (v in names(train_data)) {
  if (is.factor(train_data[[v]])) {
    test_data[[v]] <- factor(
      test_data[[v]],
      levels = levels(train_data[[v]]),
      ordered = is.ordered(train_data[[v]])
    )
  } else if (is.integer(train_data[[v]])) {
    test_data[[v]] <- as.integer(test_data[[v]])
  } else if (is.numeric(train_data[[v]])) {
    test_data[[v]] <- as.numeric(test_data[[v]])
  }
}

rf_pred_prob <- predict(rf_model, newdata = test_data, type = "prob")[, "Yes"]
rf_pred <- predict(rf_model, newdata = test_data, type = "response")

rf_cm <- confusionMatrix(rf_pred, test_data$readmit30, positive = "Yes")
rf_cm
```

## Random forest ROC-AUC

```{r}
rf_roc <- roc(test_data$readmit30, rf_pred_prob, 
              levels = c("No", "Yes"), direction = "<")

rf_auc <- auc(rf_roc)
rf_auc

plot(rf_roc, col = "grey", lwd = 2,
     main = "ROC Curve - Random Forest")
```

## Important variables from random forest

```{r}
rf_importance <- importance(rf_model)
rf_importance_table <- data.frame(
  Variable = rownames(rf_importance),
  MeanDecreaseGini = rf_importance[, "MeanDecreaseGini"],
  row.names = NULL
)
rf_importance_table <- rf_importance_table[order(-rf_importance_table$MeanDecreaseGini), ]
head(rf_importance_table, 15)
```

## Model evaluation table

```{r}
results_table <- data.frame(
  Model = c("Logistic Regression", "Random Forest"),
  Accuracy = c(log_cm$overall["Accuracy"], rf_cm$overall["Accuracy"]),
  Sensitivity = c(log_cm$byClass["Sensitivity"], rf_cm$byClass["Sensitivity"]),
  Specificity = c(log_cm$byClass["Specificity"], rf_cm$byClass["Specificity"]),
  AUC = c(as.numeric(log_auc), as.numeric(rf_auc))
)
results_table
```

## ROC comparison plot

```{r}
plot(log_roc, col = "pink", lwd = 2,
     main = "ROC Comparison: Logistic Regression vs Random Forest")

plot(rf_roc, col = "grey", lwd = 2, add = TRUE)

legend("bottomright",
       legend = c(paste("Logistic Regression AUC =", round(log_auc, 3)),
                  paste("Random Forest AUC =", round(rf_auc, 3))),
       col = c("pink", "grey"),
       lwd = 2)
```

## Discussion:
Both logistic regression and random forest models achieved high accuracy (~88%), but this was mainly due to the class imbalance, where most patients were not readmitted. Therefore, accuracy alone is not a reliable measure of model performance.

Logistic regression showed slightly better discrimination than random forest (AUC: 0.644 vs 0.617), although both models demonstrated only moderate predictive ability. A major limitation of both models was the extremely low sensitivity, meaning they failed to correctly identify most patients who were readmitted within 30 days. In contrast, specificity was very high, indicating strong performance in identifying non-readmitted patients.

Overall, while logistic regression was slightly better and more interpretable, both models are limited in practical use due to poor detection of readmitted patients. Improvements such as handling class imbalance or applying more advanced modelling techniques could enhance performance.


## Conclusion:

Logistic regression performed slightly better than random forest in terms of AUC, while both models showed similar accuracy. However, both models had very low sensitivity, indicating poor ability to detect readmitted patients.